Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-25 sections/tensors.tex
more tensors

\subsection{Level Formats}
Having described the 8 basic level structures and the functions that define a
level format, we can now detail the key level structures that we support, some
of the challenges in their implementation, and how we overcome them.

\begin{enumerate}
\item SparseList
\item SparseDict
\item ...
\end{enumerate}

>>>>

\subsection{Level Formats}
Having described the 8 basic level structures and the functions that define a
level format, we can describe the concrete level formats supported by Finch,
some of the challenges in their implementation, and how we overcome them. Note
that these formats are not meant to be exhaustive; we will later give
recommendations for additional level formats to be implemented and explain how
users can also add their own custom formats. Note that all of the levels below store their shape and a sublevel.

\begin{enumerate}
\item[Dense]
    The dense format is the simplest format, and simply maps $fiber(l, p)[i]
    \rightarrow fiber(l.lvl, p * l.shape + i)$. This format is used to store dense data,
    and is often a convenient format for the root level of a tensor. Because
    of its simplicity, freezing and thawing the level are no-ops.
\item[DenseRLE]
    The DenseRLE format is used to represent runs of repeated values. Notice
    that a challenge arises when trying to write to a level with runs: it is
    difficult to merge duplicate runs. Such a scenario might arise when merging
    runs of subfibers of length 3, representing colors in an image.  Ideally, we
    would be able to detect duplicate subfibers and merge them on the fly, but
    we cannot determine which subfibers are equal because we cannot read them
    while the sublevel is in update-only mode. Instead, we merge the duplicates
    during the freeze phase. We $freeze$ the main sublevel, $declare$ a separate
    sublevel as a buffer to store the deduplicated subfibers, and we can then
    compare each of the subfibers in the main level, copying the deduplicated
    subfibers into the buffer:
    \begin{enumerate}
        \item[$right$] A vector of indices corresponding to the end of each run in the level
        \item[$ptr$] A vector of delimiters such that $right[ptr[p]:ptr[p+1] - 1]$ is the set of delimiters in the subfiber at position $p$.
        \item[$buf$] A duplicate sublevel, used to store deduplicated subfibers during $freeze$.
    \end{enumerate}
\item[SparseList]
    The sparse list format is the simplest sparse format, and is used to
    construct the popular CSR, CSC, DCSR, DCSC, and CSF formats. The sparse list
    format consists of the following fields:
    \begin{enumerate}
        \item[$idx$] A vector of nonzero indices in the level
        \item[$ptr$] A vector of delimiters such that $idx[ptr[p]:ptr[p+1] - 1]$ is the set of nonzero indices in the subfiber at position $p$.
    \end{enumerate}
    All of Finch's sparse formats use a dirty bit during writing to determine
    whether the sublevel has been modified from it's default fill value and
    thus, whether it needs to be stored.
\item[SparsePinpoint]
    Because the SparsePinpoint format will only ever store one nonzero in each subfiber,
    we do not need the $ptr$ field.
    \begin{enumerate}
        \item[$idx$] A vector of nonzero indices in the level, one for each subfiber
    \end{enumerate}
\item[SparseRLE]
    Similar to the DenseRLE level, but because the runs are stored on top of a background of fill values, we must also store the start of each run.
    \begin{enumerate}
        \item[$left$] A vector of indices corresponding to the beginning of each run in the level
        \item[$right$] A vector of indices corresponding to the end of each run in the level
        \item[$ptr$] A vector of delimiters such that $left[ptr[p]:ptr[p+1] - 1]$ and $right[ptr[p]:ptr[p+1] - 1]$ are the set of delimiters in the subfiber at position $p$.
        \item[$buf$] A duplicate sublevel, used to store deduplicated subfibers during $freeze$.
    \end{enumerate}
\item[SparseInterval]
    Similar to the SparseRLE level, but $ptr$ is redundant because we only store
    one run. We also don't bother with deduplication here as we cannot store any
    intermediate results with duplicates.
    \begin{enumerate}
        \item[$left$] A vector of indices corresponding to the beginning of each run in the level
        \item[$right$] A vector of indices corresponding to the end of each run in the level
    \end{enumerate}
\item[SparseVBL]
    The SparseVBL format is used to represent blocked data. The format consists
    of the following fields:
    \begin{enumerate}
        \item[$idx$] A vector of indices at the end of each nonzero block in the level
        \item[$ptr$] A vector of delimiters such that $idx[ptr[p]:ptr[p+1] - 1]$ is the set of block terminals in the subfiber at position $p$.
        \item[$ofs$] A vector of delimiters such that $ofs[ptr[p] + q]:ofs[ptr[p] + q + 1] - 1$ are the subpositions of block $q$ in subfiber $p$.
    \end{enumerate}
\item[SparseBand]
    Like SparseVBL, but only one block is stored in each subfiber.
        \item[$idx$] The index at the end of the block in the level
        \item[$ofs$] A vector of delimiters such that $ofs[p]:ofs[p + 1] - 1$ are the subpositions of the block in subfiber $p$.
\end{enumerate}

\help{we need also to talk about pattern and element}

\subsection{Flexible Level Formats}
The levels we described in the previous section are all oriented towards bulk,
sequential creation of formats. However, when users want to be able to write out
of order (which is a common requirement arising from loop order or from the
problem itself, it occurs in our spgemm algorithms and our histogram example in
the evaluation section), we must use more complicated datastructures like hash
tables and trees to support the random access. Because these datastructures are
more complex and have a higher implementation burden and performance overhead,
we only support random access construction of sparse or dense structures.  We
can use these two more general structures as intermediates to convert to our
more specialized structures later.

%Future
%work might involve supporting random access construction for additional
%structures, especially those with asymptotic benefits such as SparseRLE, with
%the understanding that we can use more general structures as intermediates to
%convert to more specialized structures later.

%This leave
